---
title: "Class 12 Homework"
author: "Zainab Fatima (PID: A16880407)"
date: "2/15/25"
format: pdf
toc: TRUE
---
## Determining Sample Size and Median Expression Levels

> Question 13. Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

Answer: The sample size for each phenotype is 108 samples for A/A, 233 for A/G, and 121 for G/G. The median expression levels for each genotype is 31.25 for A/A, 25.06 for A/G, and 20.07 for G/G. The code is listed below. 

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```
How many of each sample do we have?

```{r}
table(expr$geno)
```

To find the median of each expression level, let's make a boxplot.

```{r}
boxplot(exp ~ geno, data = expr, xlab = "genotype", ylab = "expression values")
```

The function `tapply()` is used to apply a function over subsets of a vector, we can use this function to find the median value of each genotype.

```{r}
tapply(expr$exp, expr$geno, median)
```

The median values of each genotype are 31.24 for A/A, 25.06 for A/G, and 20.07 for G/G. 

## Generating Boxplot

> Q14. Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

Boxplot drawn below using ggplot. I can infer that expression value between A/A and G/G is different and that A/A SNP promotes more expression compared to G/G. This is because of the location of the "notches" in the boxplots and the medians of the graphs show that the level of A/A is higher than G/G. The SNP effects the expression of ORMDL3 based on the different levels of each group (A/A is highest expression level, A/G is medium, and G/G is the lowest expression level).  

```{r}
library(ggplot2)
graph_of_exp <- ggplot(expr) + aes(geno, exp, fill = geno) + 
  geom_boxplot(notch = TRUE) +
  labs(title="Gene Expression by Genotype",
        x ="Genotype", y = "Expression Level", fill = "Genotype")

graph_of_exp
```






