---
title: "Class 18: Pertussis Vaccination Mini Project"
author: "Zainab Fatima (PID: A16880407)"
date: "03-06-2025"
format: html
toc: true 
---
## Investigating pertussis cases by year

Pertussis (a.k.a. Whooping Cough) is a deadly lung infection caused by the bacteria B. Pertussis. 

The CDC tracks Pertussis cases around the U.S. 
http://tinyurl.com/pertussiscdc

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

We can "scrape" this data using the R **datapasta** package. 

```{r, include=FALSE}
cdc <- data.frame(
  year = c(1922L,1923L,1924L,1925L,
           1926L,1927L,1928L,1929L,1930L,1931L,
           1932L,1933L,1934L,1935L,1936L,
           1937L,1938L,1939L,1940L,1941L,1942L,
           1943L,1944L,1945L,1946L,1947L,
           1948L,1949L,1950L,1951L,1952L,
           1953L,1954L,1955L,1956L,1957L,1958L,
           1959L,1960L,1961L,1962L,1963L,
           1964L,1965L,1966L,1967L,1968L,1969L,
           1970L,1971L,1972L,1973L,1974L,
           1975L,1976L,1977L,1978L,1979L,1980L,
           1981L,1982L,1983L,1984L,1985L,
           1986L,1987L,1988L,1989L,1990L,
           1991L,1992L,1993L,1994L,1995L,1996L,
           1997L,1998L,1999L,2000L,2001L,
           2002L,2003L,2004L,2005L,2006L,2007L,
           2008L,2009L,2010L,2011L,2012L,
           2013L,2014L,2015L,2016L,2017L,2018L,
           2019L,2020L,2021L,2022L, 2024L),
  
  cases = c(107473,164191,165418,152003,
                                   202210,181411,161799,197371,
                                   166914,172559,215343,179135,265269,
                                   180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,
                                   120718,68687,45030,37129,60886,
                                   62786,31732,28295,32148,40005,
                                   14809,11468,17749,17135,13005,6799,
                                   7717,9718,4810,3285,4249,3036,
                                   3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,
                                   3589,4195,2823,3450,4157,4570,
                                   2719,4083,6586,4617,5137,7796,6564,
                                   7405,7298,7867,7580,9771,11647,
                                   25827,25616,15632,10454,13278,
                                   16858,27550,18719,48277,28639,32971,
                                   20762,17972,18975,15609,18617,
                                   6124,2116,3044, 35493)
)
```


```{r}
head(cdc)
```
## A tale of two vaccines (wP and aP)

```{r}
library(ggplot2)
```

```{r}
ggplot(cdc) + 
  aes(x = year, y = cases) +
  geom_line()  
```

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) + 
  aes(x = year, y = cases) +
  geom_line() + 
  geom_vline(xintercept = 1946, col = "blue") +
  geom_vline(xintercept = 1996, col = "red") +
  geom_vline(xintercept = 2020, col = "gray") +
  geom_vline(xintercept = 2004, col = "darkgreen")
```

There were high case numbers before the first wP (whole cell) vaccine rollout in 1946. Then there was a rapid decline in case numbers until 2004 (4 years after aP vaccine rollout) when we have our first large-scale outbreak of pertussis again. There is also a notable COVID related dip and recent rapid rise. 

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

CDC data shows a resurgence of pertussis, with 48,277 cases reported in 2012 which is the highest levels since 1955. Possible causes include improved PCR testing, vaccine hesitancy, bacterial evolution, and waning immunity from the newer aP vaccine. The aP vaccine only has the antigens of the bacteria instead of the "whole-cell" inactivated bacteria, so we can do more analysis to see what is different about the immune response to infection in patients that had the older wP vaccine compared to the newer aP vaccine.

## 3. Exploring Computational Models of Immunity Pertussis Boost (CMI-PB) Data

The CMI-PB project aims to address this key question: what is different between aP and wP individuals. 

We can get all the data from this ongoing project via JSON API calls. For this we will use the **jsonlite** package. We can install with `install.packages("jsonlite")` 

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)

head(subject)
```

How many individuals "subjects" are in this dataset? 

```{r}
nrow(subject)
```
172 individuals

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```
> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```
> Q6. What is the breakdown of race and biological sex?

```{r}
table(subject$race, subject$biological_sex)
#produces a cross-table
```
Not representative of the US population (more representative of UCSD population), so the conclusions are not going to be representative. 

## Side-Note: Working with dates

```{r}
library(lubridate)
```

What is today's date? 

```{r}
today()
```
```{r}
today() - ymd("2000-01-01")
```
What is this in years?

```{r}
time_length( today() - ymd("2000-01-01"),  "years")
```
we are using the `ymd()` function to tell lubridate the format of our particular date and then the `time_length()` function to convert days to years.

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
# Use todays date to calculate age in days
subject$age <- today() - ymd(subject$year_of_birth)
```

```{r}
library(dplyr)

ap <- subject %>% filter(infancy_vac == "aP")

round( summary( time_length( ap$age, "years" ) ) )
```

```{r}
# wP
wp <- subject %>% filter(infancy_vac == "wP")
round( summary( time_length( wp$age, "years" ) ) )
```

> Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

The two groups look different with the age of the individuals that took the aP vaccine being overall younger. We can also calculate a p-value to determine significance. 

```{r}
# Or use wilcox.test() 
x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value
```
## Joining multiple tables

Obtain more data from CMI-PB:
```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", simplifyVector = T)

ab_data <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector= T)
```

```{r}
head(specimen)
```
```{r}
head(ab_data)
```

I now have three tables of data from CMI-PB: `subject`, `specimen`, and `ab_data`. I need to "join" these tables so I will have all the info I need to work with. 

For this we will use the `inner_join()` function from **dplyr** package. 

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)

meta <- inner_join(specimen, subject)
head(meta)
```


```{r}
dim(subject)
dim(specimen)
```

```{r}
dim(meta)
```

We can join our `ab_data` table to `meta` so we have all the info we need about antibody levels.

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(meta, ab_data)
head(abdata)
```

How many different antibody isotypes are there in this dataset?

```{r}
length(abdata$isotype) #total number for isotypes
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
#How many unique isotypes?
table(abdata$isotype)
```
```{r}
#how many diff antigens?
table(abdata$antigen)
```
Measles act as a control, not expected to see anything if someone gets pertussis vaccine. More likely to see higher levels of pertussis toxin if recieving pertussis vaccine. 

> Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```
There are 4 different $dataset values, and there are 15050 rows for the most recent 2023 dataset. 

I want a plot of antigen levels across the whole dataset. 

```{r}
ggplot(abdata) +
  aes (x=MFI, y= antigen) +
  geom_boxplot()
```


```{r}
#Plot using MFI_normalised
ggplot(abdata) +
  aes (x=MFI_normalised, y= antigen) +
  geom_boxplot()
```

Antigens like FIM2/3, PT, FELD1 have a large range of values since they are the toxins in the pertussis vaccine. Others like Measles don't show much activty because they act as a control.

Q. Are the antibody levels different at the whole-dataset levels in wP vs aP individuals? 

```{r}
ggplot(abdata) +
  aes (x=MFI_normalised, y= antigen, col = infancy_vac) +
  geom_boxplot()
```


```{r}
ggplot(abdata) +
  aes (x=MFI_normalised, y= antigen, col = infancy_vac) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
#facet_wrap pulls it into two seperate columns
```

## 4. Examine IgG Ab titer levels

For this I need to select out just isotype G. 

```{r}
igg <- abdata |>
  filter(isotype == "IgG")
head(igg)
```

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens

```{r}
ggplot(igg) +
  aes (x=MFI_normalised, y= antigen, col = infancy_vac) +
  geom_boxplot()
```


```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col = infancy_vac) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

We see differences in level of IgG antibody titers recognizing antigens like FIM2/3, PT, FELD1 (have a large range of values) since they are the toxins present in the pertussis vaccines. Others like Measles don't show much activity because they are not in the vaccines and act as a control.

Digging in further to look at the time course of IgG isotype PT antigen levels across aP and wP individuals: 

```{r}

abdata.21 <- abdata %>% filter(dataset == "2021_dataset")
abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```


