---
title: "Class 10 Section 4: Comparative Structure Analysis"
author: "Zainab Fatima (PID: A16880407)"
format: pdf
date: "2-11-25"
toc: TRUE
---

Here we run through a complete "pipeline" of structure analysis that begins with a single sequence identifier and ends in a PCA analysis. 

```{r}
library(bio3d)
```

## Step 1. Get Protein Seqeucnce

Retrieve a sequence for the protein we are interested in. We will take ADK "1ake_A"

```{r}
id <- "1ake_A" 
aa <-get.seq(id)
aa
```

## Step 2. BLAST search

Run a BLAST search of the PDB for all related sequences to our input `aa`

```{r}
blast <- blast.pdb(aa)
```

```{r}
#blast$hit.tbl shows table of results

hits <- plot(blast)
#hits has details of top hits
```

## Step 3: Donwload structures 

These are our "top hits" i.e. all structures in the PDB database related to our input sequence. 

```{r}
hits$pdb.id
```

```{r}
# Download releated PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```


```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```
```{r}
pdbs
```

## Step 5: PCA Analysis

Let's use our old friend PCA to make sense of these confusing, complicated structure relationships. 

```{r}
pc <- pca(pdbs) 
plot(pc)
# shows there are 3 different clusters --> 3 different conformations
```

Let's make a trajectory (or movie) of the main conformational changes captured by PC1. We will use the `mktrj()` function for this...

```{r}
mktrj(pc, file = "pca_result.pdb")
```


Back of the envelope comparison of the PDB size to UniProt

```{r}
uniprot <- 253206171
pdb <- 231029
pdb/uniprot * 100

#why do the structures in PDB lag behind?
#strcutures are expensive and take a long time to solve 
```

