---
title: "Class 17 Lab: Analyzing sequencing data in the cloud"
author: "Zainab Fatima (PID: A16880407)"
date: "03-08-2025"
format: html
toc: TRUE
---

Instance ID 

<--- i-05180c9de5060cffb (bimm143_zainab) --->

Open an SSH client.

Locate your private key file. The key used to launch this instance is bioinf_zainab.pem

Run this command, if necessary, to ensure your key is not publicly viewable.
chmod 400 "bioinf_zainab.pem"

Connect to your instance using its Public DNS:
ec2-34-220-203-193.us-west-2.compute.amazonaws.com

## Downstream Analysis

```{r}
library(tximport)

# setup the folder and filenames to read
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```
We can calculate our estimated transcript counts for each sample in R.

```{r}
head(txi.kallisto$counts)
```
Now, we can try to see how many transcripts we have for each sample:

```{r}
colSums(txi.kallisto$counts)
```

How many transcripts are detected in at least one sample: 

```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

We need to filter out those annotated transcripts with no reads before we more onto more downstream analysis. 

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```

Let's also remove those with no change over the samples:
```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```

## Principal Component Analysis

We can do PCA of the transcriptomic profiles of the samples. In the principal components, we can scale each transcript's measured levels so they contribute equally to the PCA.

```{r}
pca <- prcomp (t(x), scale = TRUE)
```

```{r}
summary (pca)
```

```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```



> Q. Use ggplot to make a figure of P1 vs PC2, PC 1 vs PC 3, and PC 2 vs PC 3. 

```{r}
library (ggplot2)
library (ggrepel)
```

```{r}
# Make a metadata object for the samples
colData <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(colData) <- colnames(txi.kallisto$counts)
```

```{r}
#Make the data.frame for ggplot
y <- as.data.frame(pca$x)
y$condition <- as.factor(colData$condition)
```

```{r}
#PC 1 vs PC 2
ggplot(y) +
  aes(PC1, PC2, col=condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```
```{r}
#PC 1 vs PC 3
ggplot(y) +
  aes(PC1, PC3, col=condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```
```{r}
#PC 2 vs PC 3
ggplot(y) +
  aes(PC2, PC3, col=condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

